Automatically Splitting a Staged Lambda Calculus

### 2015-07-10 paper/splitting.tex
some iteration on splitting

\TODO add related work

@@

\TODO prior work statement

@@

The correctness of the splitting translation requires that the simple 
evaluate-and-plug semantics on the output produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.

\TODO defin the gap, then close it with the lemma
To help this case, we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.

From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

The above proof of \ref{thm:allCorrect} invokes \ref{lem:twoCorrect},
In order to prove that lemma, 
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

The correctness of \bbonem\ term splitting with respect to $\stepsym \bbonem$ steps is given in \ref{lem:oneCorrect}.
As in the world \bbtwo\ version, correctness requires that the applied resumer from either side of the step be equivalent.
But additionally, it also requires that first stage results be exactly identical.

@@

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:semantics}), this annotation is
particularly useful for splitting, since splitting $\pure e$
terms trivially yields a unit resumer.

An illustrative example is the \texttt{partition} function in the quickselect
example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
instead of \bbonep, then (like \texttt{qs}) it would split into two functions
\texttt{partition1} and \texttt{partition2}, the latter of which recursively
computes the (trivial) second-stage component of \texttt{partition}'s result.
Moreover, \texttt{qs2} would call \texttt{partition2}, just as \texttt{qs1}
calls \texttt{partition1}:
\begin{lstlisting}
2`fun qs2 (p : tree, k : int) : int =
... | Branch (i,h,p1,p2) =>
  let val () = partition2 ((),()) in
  case compare k i of ...
\end{lstlisting}
By explicitly marking that \texttt{partition} has no second-stage computations,
we avoid both generating the resumer \texttt{partition2}, and calling it in
\texttt{qs2}.

\TODO somebody should read the new text about partition

\end{abstrsyn}

>>>>

@@

@@

The dynamic correctness of the splitting translation requires that the simple 
evaluate-and-plug semantics on the output produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual $q$ produced by direct evaluation, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.
This condition is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
From here, the inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
All that we now require is $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
To close this gap we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.
After invocation of that new lemma, we can derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

In order to prove that \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

The correctness of \bbonem\ term splitting with respect to a $\stepsym \bbonem$ step is given in \ref{lem:oneCorrect}.
It's very similar to the world \bbtwo\ version, 
in that the reduction of the first stage part of $e'$ should imply reduction of the first stage part of $e$,
and that the resulting applied resumers should be equivalent.
But split forms at \bbonem\ have one more piece of output than those at \bbtwo,
namely the immediate value $i$.
\ref{lem:oneCorrect} accounts for this by saying that 
that immediate value must be exactly identical on both sides of the step.

@@

An illustrative example is the \texttt{partition} function in the quickselect
example of \ref{sec:exampleQS}. If \texttt{partition} were defined at \bbonem\
then (like \texttt{qs}) it would split into two functions
\texttt{partition1} and \texttt{partition2}, the latter of which recursively
computes the (trivial) second-stage component of \texttt{partition}'s result.
Moreover, \texttt{qs2} would call \texttt{partition2}, just as \texttt{qs1}
calls \texttt{partition1}:
\begin{lstlisting}
2`fun qs2 (p : tree, k : int) : int =
... | Branch (i,h,p1,p2) =>
  let val () = partition2 ((),()) in
  case compare k i of ...
\end{lstlisting}

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  
Defining \texttt{partition} in this world is 
tantamount to proving it has no second-stage computations,
allowing splitting to avoid generating the resumer \texttt{partition2}
and calling it from \texttt{qs2}.

Recall that this annotation is not especially useful 
for ordinary staged term evaluation (\ref{sec:semantics}).
We suspect the key factor is that splitting is a static transformation,
since other static trasofrmations \cite{Mogensen89a,DeNiel91} 
also benefited from a clear distinction between ground and mixed terms.
Ground is also similar to the validity mechanism in ML5 \cite{murphy05}.
\end{abstrsyn}
